The Impact of Alternation

### 2017-05-14 draft.tex
ok

The boolean operations of union, intersection and complement can be
implemented in linear time in this model, thus matching the complexity
of performing these operations in the finite-alphabet case. The price
to be paid is that emptiness becomes undecidable, for which reason we
provide two efficient semi-algorithms for emptiness, based on lazy
predicate abstraction \cite{blast} and the \impact~method
\cite{mcmillan06}.  These algorithms are proven to terminate and
return a word from the language, if the language is not empty, but
termination is not guaranteed when the language is empty.

@@

We have implemented the boolean operations and emptiness checking
semi-algorithms in a prototype tool and carried out experiments with
examples taken from array logics \cite{cav09}, timed automata
\cite{henzinger:RealTimeSystems}, communication protocols \cite{abp}
and hardware verification \cite{smrcka}.

\paragraph{Related Work}
Data languages and automata have been defined previously, in a
classical nondeterministic setting. For instance, Kaminski and Francez
\cite{Kaminski94} consider languages, over an infinite alphabet of
data, recognized by automata with a finite number of registers, that
store the input data and compare it using equality. Just as the timed
languages recognized by timed automata \cite{AlurDill94}, these
languages, called quasi-regular, are not closed under complement, but
their emptiness is decidable. The impossibility of complementation
here is caused by the use of hidden variables, which we do not
allow. Even though closed under complement, emptiness is undecidable
in our case, mainly because counting (incrementing and comparing to a
constant) data values is allowed, in many data theories.

Other approaches to defining notions of data automata
\cite{Bojanczyk:2011:TLD,BouyerPetitTherien03} are different from ours
mainly because these automata are controlled by equivalence relations
of finite index, whereas in our case, the transitions are defined by
unrestricted formulae in the first-order theory of the data domain. As
in \cite{Kaminski94,AlurDill94}, the emptiness problem is decidable
for these classes.

Another related model is that of predicate automata
\cite{Farzan15,Farzan16}, which recognize languages over integer data
by labeling the words with conjunctions of uninterpreted
predicates. We intend to explore further the connection with our model
of alternating data automata, in order to apply our method to the
verification of parallel programs.

@@

The emptiness check is implemented by adapting similar semi-algorithms
for nondeterministic infinite-state programs to the alternating model
of computation. In particular, we considered the state-of-the-art
\impact~ procedure \cite{mcmillan06} that is shown to outperform lazy
predicate abstraction in the nondeterministic case, and generalized it
to cope with alternation. More recent approaches for interpolant-based
abstraction refinement \cite{McMillan14,Hoder12} target Horn systems,
used to encode recursive and concurrent programs
\cite{Grebenshchikov12}. However, the emptiness of alternating word
automata cannot be directly encoded using Horn clauses, because all
the branches of the computation synchronize on the same input, which
does not seem to be easily encoded using a finite number of local
(equality) constraints. We believe that the lazy annotation techniques
for Horn clauses are suited for branching computations, which we
intend to consider in a future tree automata setting.

@@

We consider an infinite countable set of \emph{variables} $\vars$,
where each $x \in \vars$ has an associated sort $\sigma(x)$. Terms are
defined recursively: a term $t$ of sort $\sigma(t)=S$ is a variable $x
\in \vars$ where $\sigma(x)=S$, or $f(t_1,\ldots,t_{\#(f)})$ where
$t_1,\ldots,t_{\#(f)}$ are terms of sort $\Data$ and $\sigma(f)=S$. An
\emph{atom} is a term of sort $\Bool$ or an equality $t \teq s$
between two terms of sort $\Data$. A \emph{literal} is an atom or its
negation. A \emph{formula} is an existentially quantified combination
of atoms using disjunction $\vee$, conjunction $\wedge$ and negation
$\neg$ and we write $\phi \rightarrow \psi$ for $\neg\phi \vee \psi$.

We denote by $\fv{\sigma}{\phi}$ the set of free (not under the scope
of an existential quantifier) variables of sort $\sigma$ in $\phi$ and
write $\fv{}{\phi}$ for $\bigcup_{\sigma \in \ssorts{\sig}}
\fv{\sigma}{\phi}$. For a variable $x \in \fv{}{\phi}$ and a term $t$
such that $\sigma(t) = \sigma(x)$, let $\phi[t/x]$ be the result of
replacing each occurrence of $x$ by $t$. For indexed sets
$\vec{t}=\set{t_1,\ldots,t_n}$ and $\vec{x}=\set{x_1,\ldots,x_n}$, we
write $\phi[\vec{t}/\vec{x}]$ for the formula obtained by
simultaneously replacing $x_i$ with $t_i$ in $\phi$, for all
$i\in[1,n]$. The size $\len{\phi}$ is the number of symbols occuring
in $\phi$.

An \emph{interpretation} $\I$ maps\begin{inparaenum}[(1)]
\item the sort $\Data$ into a non-empty set $\Data^\I$, 
%
\item the sort $\Bool$ into the set $\booli = \set{\true,\false}$, where
  $\top^\I = \true$, $\bot^\I = \false$, and
%
\item each function symbol $f$ into a total function $f^\I :
  (\Data^\I)^{\#(f)} \rightarrow \sigma(f)^I$, or an element of
  $\sigma(f)^I$ when $\#(f)=0$.
\end{inparaenum}
Given an interpretation $\I$, a \emph{valuation} $\nu$ maps each
variable $x \in \vars$ into an element $\nu(x) \in \sigma(x)^\I$. For
a term $t$, we denote by $t^\I_\nu$ the value obtained by replacing
each function symbol $f$ by its interpretation $f^\I$ and each
variable $x$ by its valuation $\nu(x)$. For a formula $\phi$, we write
$\I,\nu \models \phi$ if the formula obtained by replacing each term
$t$ in $\phi$ by the value $t^\I_\nu$ is logically equivalent to true.
A formula $\phi$ is \emph{satisfiable} in the interpretation $\I$ if
there exists a valuation $\nu$ such that $\I,\nu \models \phi$, and
\emph{valid} if $\I,\nu \models \phi$ for all valuations $\nu$.  The
\emph{theory} $\theory(\sig,\I)$ is the set of valid formulae written
in the signature $\sig$, with the interpretation $\I$. A decision
procedure for $\theory(\sig,\I)$ is an algorithm that takes a formula
$\phi$ in the signature $\sig$ and returns yes iff $\phi \in
\theory(\sig,\I)$.

@@

An \emph{alternating data automaton} (ADA or automaton in the sequel)
is a tuple $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$,
where: \begin{compactitem}
%
\item $\vec{x} \subset \vars$ is a finite set of variables of sort
  $\Data$,
%
\item $Q$ is a finite set of variables of sort $\Bool$, called
  \emph{states},
%
\item $\iota \in \Form^+(Q,\emptyset)$ is positive boolean combination
  of states defining the \emph{initial configurations},
%
\item $F \subseteq Q$ is a set of \emph{final states}, and
%
\item $\Delta : Q \times \Sigma \rightarrow
  \Form^+(Q,\overline{\vec{x}}\cup\vec{x})$ is a \emph{transition
    function}, where $\overline{\vec{x}}$ is the set $\{\overline{x}
  \mid x \in \vec{x}\}$.
\end{compactitem}
Intuitively, in each formula $\Delta(q,a)$ describing a transition
rule, the variables $\overline{\vec{x}}$ track the previous and
$\vec{x}$ the current values of the variables of $\A$. Observe
further, that the initial values of the variables are left
unconstrained, as the initial configuration does not contain free
data variables. The size of $\A$ is defined as $\len{\A} = \len{\iota}
+ \sum_{(q,a) \in Q\times\Sigma}\len{\Delta(q,a)}$.

@@

\paragraph{\em Example}
Figure \ref{fig:ex} depicts an alternating data automaton, with input
alphabet $\Sigma = \set{a,b}$, variables $\vec{x} = \set{x,y}$, states
$Q = \set{q_0,q_1,q_2,q_3,q_4}$, initial configuration $q_0$, final
states $F=\set{q_3,q_4}$ and transitions:
\[\begin{array}{rcl}
\Delta(q_0,a) & \equiv & q_1 \wedge q_2 \wedge x=0 \wedge y=0 \\
\Delta(q_1,a) & \equiv & q_1 \wedge q_3 \wedge x=\overline{y}+1 \wedge y=\overline{x}+1 \\
\Delta(q_1,b) & \equiv & q_3 \wedge \overline{x} \geq \overline{y} \\
\Delta(q_2,a) & \equiv & q_2 \wedge x > \overline{x} \wedge y > \overline{y} \\
\Delta(q_2,b) & \equiv & q_4 \wedge \overline{x}>\overline{y}
\end{array}\]
The missing transition rules are assumed to be $\false$. Rules
$\Delta(q_0,a)$ and $\Delta(q_1,a)$ are universal and there are no
existential nondeterministic rules. Rules $\Delta(q_1,a)$ and
$\Delta(q_2,b)$ compare past ($\overline{x},\overline{y}$) with present
($x,y$) values, $\Delta(q_0,a)$ constrains the present and
$\Delta(q_1,b)$, $\Delta(q_2,b)$ the past values, respectively. \qed

Formally, let $\vec{x}_k = \set{x_k \mid x \in \vec{x}}$, $k\geq0$, be
a set of time-stamped variables. For an input event $a \in \Sigma$ and
a formula $\phi$, we write $\Delta(\phi,a)$ (respectively,
$\Delta^k(\phi,a)$) for the formula obtained from $\phi$ by
simultaneously replacing each state $q \in \fv{\Bool}{\phi}$ by the
formula $\Delta(q,a)$ (respectively
$\Delta(q,a)[\vec{x}_k/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]$, for
all $k\geq0$). Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots
(a_n,\nu_n)$, the \emph{run} of $\A$ over $w$ is the sequence of
formulae $\phi_0,\phi_1,\ldots,\phi_n$, where $\phi_0 \equiv \iota$
and, for all $k\in[1,n]$, we have $\phi_k \equiv
\Delta^k(\phi_{k-1},a_k)$. Next, we slightly abuse notation and write
$\Delta(\iota,a_1,\ldots,a_n)$ for the formula
$\phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ above. We say that $\A$
\emph{accepts} $w$ iff $\I,\nu \models \Delta(\iota,a_1,\ldots,a_n)$,
where $\nu$ is some valuation that maps:\begin{inparaenum}[(1)]
%
\item each $x \in \vec{x}_k$ to $\nu_k(x)$, for all $k\in[1,n]$, 
%
\item each $q \in \fv{\Bool}{\phi_n} \cap F$ to $\top$ and 
%
\item each $q \in \fv{\Bool}{\phi_n} \setminus F$ to $\bot$.
\end{inparaenum}
The language of $\A$ is the set $L(\A)$ of words from
$\Sigma[\vec{x}]^*$ accepted by $\A$.

@@

Given a set $Q$ of boolean variables and a set $\vec{x}$ of variables
of sort $\Data$, for a formula $\phi \in \Form^+(Q,\vec{x})$, with no
negated occurrences of the boolean variables, we define the formula
$\overline{\phi} \in \Form^+(Q,\vec{x})$ recursively on the structure
of $\phi$:
\[\begin{array}{lclclcl}
\overline{\phi_1 \vee \phi_2} & \equiv & \overline{\phi_1} \wedge \overline{\phi_2} && 
\overline{\phi_1 \wedge \phi_2} & \equiv & \overline{\phi_1} \vee \overline{\phi_2} \\ 
\overline{\neg\phi} & \equiv & \neg \overline{\phi} \text{ if $\phi$ not atom} &&
\overline{\phi} & \equiv & \phi \text{ if $\phi \in Q$} \\
\overline{\phi} & \equiv & \neg\phi \text{ if $\phi \not\in Q$ atom}
\end{array}\]
It is easy to check that $\len{\overline{\phi}} = \len{\phi}$, for
every formula $\phi \in \Form^+(Q,\vec{x})$.

@@

It is easy to see that $\len{\A_\cup} = \len{\A_\cap} = \len{\A_1} +
\len{\A_2}$ and $\len{\overline{\A}} = \len{\A}$, thus the automata
for the boolean operations, including complementation, can be built in
linear time. This is not surprising, as linear time intersection and
complementation are also possible for alternating automata over finite
alphabets \cite{ChandraKozenStockmeyer81}.

@@

We generalize the post-image function to finite sequences of input
events, as follows, for any $u \in \Sigma^*$:
\[\begin{array}{l}
\Post{\A}(\phi,\varepsilon) \equiv \phi \hspace*{5mm} \Post{\A}(\phi,ua) \equiv \Post{\A}(\Post{\A}(\phi,u),a) \\
\Accept{\A}(u) \equiv \Post{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} (q \rightarrow \bot)
\end{array}\]
Then the emptiness problem for $\A$ becomes: does there exist $u \in
\Sigma^*$ such that the formula $\Accept{\A}(u)$ is satisfiable?
Observe that, since we ask a satisfiability query, the final states of
$\A$ need not be constrained\footnote{ Since each state occurs
  positively in $\Accept{\A}(u)$, this formula has a model iff it has
  a model with every $q \in F$ set to true.}. A na\"ive semi-algorithm
enumerates all finite sequences and checks the satisfiability of
$\Accept{\A}(u)$ for each $u \in \Sigma^*$, using a decision procedure
for the theory $\theory(\sig,\I)$.

@@

Formally, we fix a set of formulae $\preds \subseteq
\Form(Q,\vec{x})$, such that $\bot \in \preds$ and refer to these
formulae as \emph{predicates}. Given a formula $\phi$, we denote by
$\abs{\phi} \equiv \bigwedge \set{\pi \in \preds \mid \phi \models
  \pi}$ the abstraction of $\phi$ w.r.t. the predicates in
$\preds$. The abstract versions of the post-image and acceptance
condition are defined as follows, for any $u \in \Sigma^*$:
\[\begin{array}{l}
\AbsPost{\A}(\phi,\varepsilon) \equiv \phi \hspace*{5mm}
\AbsPost{\A}(\phi,ua) \equiv \abs{(\Post{\A}(\AbsPost{\A}(\phi,u),a))} \\
\AbsAccept{\A}(u) \equiv \AbsPost{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} 
(q \rightarrow \bot)
\end{array}\]

\begin{lemma}\label{lemma:abstract-invariant}
  For any bijection $\mu : \nat \rightarrow \Sigma^*$, there exists
  $k>0$ such that $\bigvee_{i=0}^k \AbsPost{\A}(\iota,\mu(i))$ defines
  an invariant for $\A$. 
\end{lemma}

@@

Given a non-empty spurious counterexample $u = a_1\ldots a_n$, where
$n>0$, we consider the following interpolation problem: 
\begin{eqnarray}\label{eq:interpolation-problem}
\Theta(u) & \equiv & \theta_0(Q_0) \wedge \theta_1(Q_0 \cup Q_1,\vec{x}_0
\cup \vec{x}_1) \wedge \ldots \\ 
&& \wedge~ \theta_n(Q_{n-1} \cup Q_n,\vec{x}_{n-1} \cup \vec{x}_n) \wedge \theta_{n+1}(Q_n) \nonumber
\end{eqnarray}
where $Q_k = \set{q_k \mid q \in Q}$, $k \in [0,n]$ are time-stamped
sets of boolean variables corresponding to the set $Q$ of states of
$\A$. The first conjunct $\theta_0(Q_0) \equiv \iota[Q_0/Q]$ is the
initial configuration of $\A$, with every $q \in \fv{\Bool}{\iota}$
replaced by $q_0$. The definition of $\theta_k$, $k\in[1,n]$ uses
\emph{replacement sets} $R_\ell \subseteq Q_\ell$, $\ell\in [0,n]$,
which are defined inductively below: \begin{compactitem}
\item $R_0 = \fv{\Bool}{\theta_0}$ and 
%
\item $\theta_\ell \equiv \bigwedge_{q_{\ell-1}\in R_{\ell-1}}
  (q_{\ell-1} \rightarrow
  \Delta(q,a_\ell)[Q_\ell/Q,\vec{x}_{\ell-1}/\overline{\vec{x}},\vec{x}_\ell/\vec{x}])$
  and $R_\ell = \fv{\Bool}{\theta_\ell} \cap Q_\ell$, for each
  $\ell\in[1,n]$.
\end{compactitem}
Finally, we set $\theta_{n+1}(Q_n) \equiv \bigwedge_{q \in Q \setminus
  F} (q_n \rightarrow \bot)$. The intuition is that $\theta_0, \ldots,
\theta_n$ encode the run of $\A$ over $u$ and $\theta_{n+1}$ is the
acceptance condition, which forces the last remaining non-final states
to be false.

@@

\begin{proposition}\label{prop:positive-interpol}
  For any $u = a_1 \ldots a_n \in \Sigma^*$, if $\pi(u)$ is
  unsatisfiable then one can build an interpolant
  $\tuple{\top,I_0,\ldots,I_n,\bot}$ for $\pi(u)$, such that $I_i \in
  \Form^+(Q,\vec{x})$, for all $i \in [0,n]$.
\end{proposition}

We recall that a run of $\A$ over $u$ is a sequence $\phi_0(Q),
\phi_1(Q,\vec{x}_0\cup\vec{x}_1), \ldots,
\phi_n(Q,\vec{x}_0\cup\ldots\cup\vec{x}_n)$, where $\phi_0$ is the
initial configuration $\iota$ and for each $k>0$, $\phi_k$ is obtained
from $\phi_{k-1}$ by replacing each state $q \in
\fv{\Bool}{\phi_{k-1}}$ by the formula
$\Delta(q,a_k)[\vec{x}_{k-1}/\overline{\vec{x}},\vec{x}_k/\vec{x}]$,
given by the transition function of $\A$. Observe that, because the
states are replaced with transition formulae when moving one step in a
run, the formulae thus produced lose track of the control history and
are not directly suitable to producing interpolants that relates
states and data.

>>>>

The boolean operations of union, intersection and complement can be
implemented in linear time in this model, thus matching the complexity
of performing these operations in the finite-alphabet case. The price
to be paid is that emptiness becomes undecidable, for which reason we
provide two efficient semi-algorithms for emptiness, based on lazy
predicate abstraction \cite{blast} and the \impact~method
\cite{mcmillan06}.  These algorithms are proven to terminate and
return a word from the language of the automaton, if one exists, but
termination is not guaranteed when the language is empty.

@@

We have implemented the boolean operations and emptiness checking
semi-algorithms and carried out experiments with examples taken from
array logics \cite{cav09}, timed automata
\cite{henzinger:RealTimeSystems}, communication protocols \cite{abp}
and hardware verification \cite{smrcka}.

\paragraph{Related Work}
Data languages and automata have been defined previously, in a
classical nondeterministic setting. For instance, Kaminski and Francez
\cite{Kaminski94} consider languages, over an infinite alphabet of
data, recognized by automata with a finite number of registers, that
store the input data and compare it using equality. Just as the timed
languages recognized by timed automata \cite{AlurDill94}, these
languages, called quasi-regular, are not closed under complement, but
their emptiness is decidable. The impossibility of complementation
here is caused by the use of hidden variables, which we do not
allow. Emptiness is however undecidable in our case, mainly because
counting (incrementing and comparing to a constant) data values is
allowed, in many data theories.

%% Other approaches to defining classes of data automata
%% \cite{Bojanczyk:2011:TLD,BouyerPetitTherien03} are different from ours
%% mainly because these automata are controlled by equivalence relations
%% of finite index, whereas in our case, the transitions are defined by
%% unrestricted formulae in the first-order theory of the data domain. As
%% in \cite{Kaminski94,AlurDill94}, the emptiness problem is decidable
%% for these classes.

Another related model is that of predicate automata \cite{Farzan15},
which recognize languages over integer data by labeling the words with
conjunctions of uninterpreted predicates. We intend to explore further
the connection with our model of alternating data automata, in order
to apply our method to the verification of parallel programs.

@@

The emptiness check for alternating data automata adapts similar
semi-algorithms for nondeterministic infinite-state programs to the
alternating model of computation. In particular, we considered the
state-of-the-art \impact~ procedure \cite{mcmillan06} that is shown to
outperform lazy predicate abstraction \cite{blast} in the
nondeterministic case, and generalized it to cope with
alternation. More recent approaches for interpolant-based abstraction
refinement target Horn systems \cite{McMillan14,Hoder12}, used to
encode recursive and concurrent programs
\cite{Grebenshchikov12}. However, the emptiness of alternating word
automata cannot be directly encoded using Horn clauses, because all
the branches of the computation synchronize on the same input, which
cannot be encoded by a finite number of local (equality)
constraints. We believe that the lazy annotation techniques for Horn
clauses are suited for branching computations, which we intend to
consider in a future tree automata setting.

@@

We consider an infinite countable set of \emph{variables} $\vars$,
where each $x \in \vars$ has an associated sort $\sigma(x)$. A
\emph{term} $t$ of sort $\sigma(t)=S$ is a variable $x \in \vars$
where $\sigma(x)=S$, or $f(t_1,\ldots,t_{\#(f)})$ where
$t_1,\ldots,t_{\#(f)}$ are terms of sort $\Data$ and $\sigma(f)=S$. An
\emph{atom} is a term of sort $\Bool$ or an equality $t \teq s$
between two terms of sort $\Data$. A \emph{formula} is an
existentially quantified combination of atoms using disjunction
$\vee$, conjunction $\wedge$ and negation $\neg$ and we write $\phi
\rightarrow \psi$ for $\neg\phi \vee \psi$.

We denote by $\fv{\sigma}{\phi}$ the set of free variables of sort
$\sigma$ in $\phi$ and write $\fv{}{\phi}$ for $\bigcup_{\sigma \in
  \ssorts{\sig}} \fv{\sigma}{\phi}$. For a variable $x \in
\fv{}{\phi}$ and a term $t$ such that $\sigma(t) = \sigma(x)$, let
$\phi[t/x]$ be the result of replacing each occurrence of $x$ by
$t$. For indexed sets $\vec{t}=\set{t_1,\ldots,t_n}$ and
$\vec{x}=\set{x_1,\ldots,x_n}$, we write $\phi[\vec{t}/\vec{x}]$ for
the formula obtained by simultaneously replacing $x_i$ with $t_i$ in
$\phi$, for all $i\in[1,n]$. The size $\len{\phi}$ is the number of
symbols occuring in $\phi$.

An \emph{interpretation} $\I$ maps\begin{inparaenum}[(1)]
\item the sort $\Data$ into a non-empty set $\Data^\I$, 
%
\item the sort $\Bool$ into the set $\booli = \set{\true,\false}$, where
  $\top^\I = \true$, $\bot^\I = \false$, and
%
\item each function symbol $f$ into a total function $f^\I :
  (\Data^\I)^{\#(f)} \rightarrow \sigma(f)^I$, or an element of
  $\sigma(f)^I$ when $\#(f)=0$.
\end{inparaenum}
Given an interpretation $\I$, a \emph{valuation} $\nu$ maps each
variable $x \in \vars$ into an element $\nu(x) \in \sigma(x)^\I$. For
a term $t$, we denote by $t^\I_\nu$ the value obtained by replacing
each function symbol $f$ by its interpretation $f^\I$ and each
variable $x$ by its valuation $\nu(x)$. For a formula $\phi$, we write
$\I,\nu \models \phi$ if the formula obtained by replacing each term
$t$ in $\phi$ by the value $t^\I_\nu$ is logically equivalent to true.

A formula $\phi$ is \emph{satisfiable} in the interpretation $\I$ if
there exists a valuation $\nu$ such that $\I,\nu \models \phi$, and
\emph{valid} if $\I,\nu \models \phi$ for all valuations $\nu$.  The
\emph{theory} $\theory(\sig,\I)$ is the set of valid formulae written
in the signature $\sig$, with the interpretation $\I$. A decision
procedure for $\theory(\sig,\I)$ is an algorithm that takes a formula
$\phi$ in the signature $\sig$ and returns yes iff $\phi \in
\theory(\sig,\I)$.

@@

An \emph{alternating data automaton} (ADA or automaton in the sequel)
is a tuple $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$,
where: \begin{compactitem}
%
\item $\vec{x} \subset \vars$ is a finite set of variables of sort
  $\Data$,
%
\item $Q$ is a finite set of variables of sort $\Bool$, called
  \emph{states},
%
\item $\iota \in \Form^+(Q,\emptyset)$ is positive boolean combination
  of states defining the \emph{initial configurations},
%
\item $F \subseteq Q$ is a set of \emph{final states}, and
%
\item $\Delta : Q \times \Sigma \rightarrow
  \Form^+(Q,\overline{\vec{x}}\cup\vec{x})$ is a \emph{transition
    function},
\end{compactitem}
where $\overline{\vec{x}}=\{\overline{x} \mid x \in \vec{x}\}$. In
each formula $\Delta(q,a)$ describing a transition rule, the variables
$\overline{\vec{x}}$ track the previous and $\vec{x}$ the current
values of the variables of $\A$. Observe that the initial values of
the variables are left unconstrained, as the initial configuration
does not contain free data variables. The size of $\A$ is defined as
$\len{\A} = \len{\iota} + \sum_{(q,a) \in
  Q\times\Sigma}\len{\Delta(q,a)}$.

@@

\paragraph{\em Example}
Figure \ref{fig:ex} depicts an ADA with input alphabet $\Sigma =
\set{a,b}$, variables $\vec{x} = \set{x,y}$, states $Q =
\set{q_0,q_1,q_2,q_3,q_4}$, initial configuration $q_0$, final states
$F=\set{q_3,q_4}$ and transitions:
\[\begin{array}{rcl}
\Delta(q_0,a) & \equiv & q_1 \wedge q_2 \wedge x=0 \wedge y=0 \\
\Delta(q_1,a) & \equiv & q_1 \wedge q_3 \wedge x=\overline{y}+1 \wedge y=\overline{x}+1 \\
\Delta(q_1,b) & \equiv & q_3 \wedge \overline{x} \geq \overline{y} \\
\Delta(q_2,a) & \equiv & q_2 \wedge x > \overline{x} \wedge y > \overline{y} \\
\Delta(q_2,b) & \equiv & q_4 \wedge \overline{x}>\overline{y}
\end{array}\]
The missing rules, such as $\Delta(q_0,b)$, are assumed to be
$\bot$. Rules $\Delta(q_0,a)$ and $\Delta(q_1,a)$ are universal and
there are no existential nondeterministic rules. Rules $\Delta(q_1,a)$
and $\Delta(q_2,a)$ compare past ($\overline{x},\overline{y}$) with
present ($x,y$) values, $\Delta(q_0,a)$ constrains the present and
$\Delta(q_1,b)$, $\Delta(q_2,b)$ the past values, respectively. \qed

Formally, let $\vec{x}_k = \set{x_k \mid x \in \vec{x}}$, $k\geq0$, be
a set of time-stamped variables. For an input event $a \in \Sigma$ and
a formula $\phi$, we write $\Delta(\phi,a)$ (respectively
$\Delta^k(\phi,a)$) for the formula obtained from $\phi$ by
simultaneously replacing each state $q \in \fv{\Bool}{\phi}$ by the
formula $\Delta(q,a)$ (respectively
$\Delta(q,a)[\vec{x}_k/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]$, for
all $k\geq0$). Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots
(a_n,\nu_n)$, the \emph{run} of $\A$ over $w$ is the sequence of
formulae: \[\phi_0(Q) \Rightarrow \phi_1(Q,\vec{x}_0 \cup \vec{x}_1)
\Rightarrow \ldots \Rightarrow \phi_n(Q,\vec{x}_0 \cup \ldots
\vec{x}_n)\] where $\phi_0 \equiv \iota$ and, for all $k\in[1,n]$, we
have $\phi_k \equiv \Delta^k(\phi_{k-1},a_k)$. Next, we slightly abuse
notation and write $\Delta(\iota,a_1,\ldots,a_n)$ for the formula
$\phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ above. We say that $\A$
\emph{accepts} $w$ iff $\I,\nu \models \Delta(\iota,a_1,\ldots,a_n)$,
where $\nu$ is some valuation that maps:\begin{inparaenum}[(1)]
%
\item each $x \in \vec{x}_k$ to $\nu_k(x)$, for all $k\in[1,n]$, 
%
\item each $q \in \fv{\Bool}{\phi_n} \cap F$ to $\top$ and 
%
\item each $q \in \fv{\Bool}{\phi_n} \setminus F$ to $\bot$.
\end{inparaenum}
The language of $\A$ is the set $L(\A)$ of words from
$\Sigma[\vec{x}]^*$ accepted by $\A$.

@@

Given a set $Q$ of boolean variables and a set $\vec{x}$ of variables
of sort $\Data$, for a formula $\phi \in \Form^+(Q,\vec{x})$, with no
negated occurrences of the boolean variables, we define the formula
$\overline{\phi} \in \Form^+(Q,\vec{x})$ recursively on the structure
of $\phi$:
\[\begin{array}{lclclcl}
\overline{\phi_1 \vee \phi_2} & \equiv & \overline{\phi_1} \wedge \overline{\phi_2} && 
\overline{\phi_1 \wedge \phi_2} & \equiv & \overline{\phi_1} \vee \overline{\phi_2} \\ 
\overline{\neg\phi} & \equiv & \neg \overline{\phi} \text{ if $\phi$ not atom} &&
\overline{\phi} & \equiv & \phi \text{ if $\phi \in Q$} \\
\overline{\phi} & \equiv & \neg\phi \text{ if $\phi \not\in Q$ atom}
\end{array}\]
We have $\len{\overline{\phi}} = \len{\phi}$, for every formula $\phi
\in \Form^+(Q,\vec{x})$.

@@

It is easy to see that $\len{\A_\cup} = \len{\A_\cap} = \len{\A_1} +
\len{\A_2}$ and $\len{\overline{\A}} = \len{\A}$, thus the automata
for the boolean operations, including complementation, can be built in
linear time. This matches the linear-time bounds for intersection and
complementation of alternating automata over finite alphabets
\cite{ChandraKozenStockmeyer81}.

@@

We generalize the post-image function to finite sequences of input
events, as follows:
\[\begin{array}{l}
\Post{\A}(\phi,\varepsilon) \equiv \phi \hspace*{5mm} \Post{\A}(\phi,ua) \equiv \Post{\A}(\Post{\A}(\phi,u),a) \\
\Accept{\A}(u) \equiv \Post{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} (q \rightarrow \bot)\text{, for any $u \in \Sigma^*$}
\end{array}\]
Then the emptiness problem for $\A$ becomes: does there exist $u \in
\Sigma^*$ such that the formula $\Accept{\A}(u)$ is satisfiable?
Observe that, since we ask a satisfiability query, the final states of
$\A$ need not be constrained\footnote{ Since each state occurs
  positively in $\Accept{\A}(u)$, this formula has a model iff it has
  a model with every $q \in F$ set to true.}. A na\"ive semi-algorithm
enumerates all finite sequences and checks the satisfiability of
$\Accept{\A}(u)$ for each $u \in \Sigma^*$, using a decision procedure
for the theory $\theory(\sig,\I)$.

@@

Formally, we fix a set of formulae $\preds \subseteq
\Form(Q,\vec{x})$, such that $\bot \in \preds$ and refer to these
formulae as \emph{predicates}. Given a formula $\phi$, we denote by
$\abs{\phi} \equiv \bigwedge \set{\pi \in \preds \mid \phi \models
  \pi}$ the abstraction of $\phi$ w.r.t. the predicates in
$\preds$. The abstract versions of the post-image and acceptance
condition are defined as follows:
\[\begin{array}{l}
\AbsPost{\A}(\phi,\varepsilon) \equiv \phi \hspace*{5mm}
\AbsPost{\A}(\phi,ua) \equiv \abs{(\Post{\A}(\AbsPost{\A}(\phi,u),a))} \\
\AbsAccept{\A}(u) \equiv \AbsPost{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} 
(q \rightarrow \bot)\text{, for any $u \in \Sigma^*$}
\end{array}\]

\begin{lemma}\label{lemma:abstract-invariant}
  For any bijection $\mu : \nat \rightarrow \Sigma^*$, there exists
  $k>0$ such that $\bigvee_{i=0}^k \AbsPost{\A}(\iota,\mu(i))$ defines
  an invariant $\abs{\Inv}$ for $\A$. 
\end{lemma}

@@

Given a non-empty spurious counterexample $u = a_1\ldots a_n$, where
$n>0$, we consider the following interpolation problem: 
\begin{eqnarray}\label{eq:interpolation-problem}
\Theta(u) & \equiv & \theta_0(Q_0) \wedge \theta_1(Q_0 \cup Q_1,\vec{x}_0
\cup \vec{x}_1) \wedge \ldots \\ 
&& \wedge~ \theta_n(Q_{n-1} \cup Q_n,\vec{x}_{n-1} \cup \vec{x}_n) \wedge \theta_{n+1}(Q_n) \nonumber
\end{eqnarray}
where $Q_k = \set{q_k \mid q \in Q}$, $k \in [0,n]$ are time-stamped
sets of boolean variables corresponding to the set $Q$ of states of
$\A$. The first conjunct $\theta_0(Q_0) \equiv \iota[Q_0/Q]$ is the
initial configuration of $\A$, with every $q \in \fv{\Bool}{\iota}$
replaced by $q_0$. The definition of $\theta_k$, for all $k\in[1,n]$,
uses \emph{replacement sets} $R_\ell \subseteq Q_\ell$, $\ell\in
[0,n]$, which are defined inductively below: \begin{compactitem}
\item $R_0 = \fv{\Bool}{\theta_0}$,
%
\item $\theta_\ell \equiv \bigwedge_{q_{\ell-1}\in R_{\ell-1}}
  (q_{\ell-1} \rightarrow
  \Delta(q,a_\ell)[Q_\ell/Q,\vec{x}_{\ell-1}/\overline{\vec{x}},\vec{x}_\ell/\vec{x}])$
  and $R_\ell = \fv{\Bool}{\theta_\ell} \cap Q_\ell$, for each
  $\ell\in[1,n]$.
%
\item $\theta_{n+1}(Q_n) \equiv \bigwedge_{q \in Q \setminus F} (q_n
  \rightarrow \bot)$.
\end{compactitem}
The intuition is that $R_0,\ldots,R_n$ are the sets of states replaced,
$\theta_0, \ldots, \theta_n$ are the transition rules fired on the run
of $\A$ over $u$ and $\theta_{n+1}$ is the acceptance condition, which
forces the last remaining non-final states to be false.

@@

\begin{proposition}\label{prop:positive-interpol}
  If $\Theta(a_1 \ldots a_n)$ is unsatisfiable then one can build an
  interpolant $\tuple{\top,I_0,\ldots,I_n,\bot}$ for $\Theta(a_1\ldots
  a_n)$, such that $I_i \in \Form^+(Q,\vec{x})$, for all $i \in
  [0,n]$.
\end{proposition}

We recall that a run of $\A$ over $u$ is a sequence: \[\phi_0(Q)
\Rightarrow \phi_1(Q,\vec{x}_0\cup\vec{x}_1) \Rightarrow \ldots
\Rightarrow \phi_n(Q,\vec{x}_0\cup\ldots\cup\vec{x}_n)\] where
$\phi_0$ is the initial configuration $\iota$ and for each $k>0$,
$\phi_k$ is obtained from $\phi_{k-1}$ by replacing each state $q \in
\fv{\Bool}{\phi_{k-1}}$ by the formula
$\Delta(q,a_k)[\vec{x}_{k-1}/\overline{\vec{x}},\vec{x}_k/\vec{x}]$,
given by the transition function of $\A$. Observe that, because the
states are replaced with transition formulae when moving one step in a
run, the formulae thus produced lose track of the control history and
are not directly suitable to producing interpolants that relate states
and data.
